---
title: "Directed Studies - Assignment1"
author: "Taruque Mahmud - T00685251"
date: "9/26/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

In this work, I have taken the dataset for breast cancer detection to develop a model of logistic regression and calculated the glm and mle. For this exploration, I have taken one variable which is the radius_mean as the dependent variable and the diagnosis as independent variable.

I have used optimx package to estimate the parameters for this dataset using Maximum Likelihood method.


#Loading packages and dataset

```{r}
require(optimx)

setwd("C:/All Files/TRU Study/2. Directed Studies/A1/")
data <- read.csv("data.csv")


class(data)
df <- as.data.frame(data)

```

Here, In this dataset, the variables are as below: 
diagnosis: M = malignant, B = benign
radius: distances from center to points on the perimeter


Now converting the diagnosis values to binary.

```{r}
df$diagnosis_n <- ifelse(df$diagnosis == "M", 1, 0)
```

Generalized linear model

```{r}

modelcp<- glm(df$diagnosis_n~df$radius_mean,family=binomial)
(result<-summary(modelcp))

plot(modelcp)

```

optim function MLE

```{r}

f1<-function(para){
  eta<-para[1]+para[2]*df$radius_mean
  p<-1/(1+exp(-eta))
  -sum(log(choose(1,df$diagnosis_n))+df$diagnosis_n*log(p)+(1-df$diagnosis_n)*log(1-p),na.rm=TRUE)
}
f2<-(optim1<-optim(c(1,1),fn=f1,hessian=TRUE))
f2
```
Plotting the functions

```{r}
#plot optim
plot(df$radius_mean,                           
     df$diagnosis_n,
     main = "optim Function")

#plot glm
plot(df$radius_mean,                           
     df$diagnosis_n,
     main = "glm Function")

```

newton-raphson method(NRM):

Equation:

```{r}

X = model.matrix(modelcp)
y = modelcp$y

```

Creating function for the Newton Raphson method

```{r}
newton <- function(
  X,
  y,
  tol  = 1e-12,
  iter = 100,
  stepsize = .5
) {

  int     = log(mean(y) / (1 - mean(y)))         # intercept
  beta    = c(int, rep(0, ncol(X) - 1))
  currtol = 1
  it = 0
  ll = 0
  
  while (currtol > tol && it < iter) {
    it = it +1
    ll_old = ll
    
    mu = plogis(X %*% beta)[,1]
    g  = crossprod(X, mu-y)               # gradient
    S  = diag(mu*(1-mu)) 
    H  = t(X) %*% S %*% X                 # hessian
    beta = beta - stepsize * solve(H) %*% g
    
    ll = sum(dbinom(y, prob = mu, size = 1, log = TRUE))
    currtol = abs(ll - ll_old)
  }
  
  list(
    beta = beta,
    iter = it,
    tol  = currtol,
    loglik = ll
  )
}


```

NRM results:

```{r}
newton_result = newton(
  X = X,
  y = y,
  stepsize = .9,
  tol = 1e-8
) 

```

Comparing the results with glm and newton raphson method

```{r}
rbind(
  newton = unlist(newton_result),
  glm_default = c(
    beta = coef(modelcp),
    modelcp$iter,
    tol = NA,
    loglik = -logLik(modelcp)
  ),
  mle = c( 
    beta = f2$par,
    f2$iter,
    tol = NA)
)
```

The resutl looks similar after 10 iteration in NRM.




